---
title: "Own Project - Predicting Peptide Retention Times"
output: html_document
---

# 1. Introduction/overview/executive summary

Chromatography is used in proteomics experiments to slowly elute different
peptide species for the subsequent mass spectrometric analysis. Goal of the here
described study is to predict the retention time (RT) of peptides based only on 
their amino acid sequence. The data set containing peptide sequences and RT was 
downloaded from:

https://www.kaggle.com/kirillpe/proteomics-retention-time-prediction

The data is also included in the uploaded files (unmod.txt).

General information about proteomics:

https://en.wikipedia.org/wiki/Proteomics

# 2. Methods/analysis

The provided peptide sequences were derived by digestion of proteins and the 
amino acid sequences of the peptides can be used to predict the physico-chemical
properties of the peptides.
General information about amino acids can be found here:

https://en.wikipedia.org/wiki/Amino_acid

The R package "Peptides" is used to predict physico-chemical properties of all
peptides.

Manual of the Peptides R package:

https://cran.r-project.org/web/packages/Peptides/Peptides.pdf

```{r}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(Peptides)) install.packages("Peptides", repos = "http://cran.us.r-project.org")
if(!require(ggpubr)) install.packages("ggpubr", repos = "http://cran.us.r-project.org")
if(!require(broom)) install.packages("broom", repos = "http://cran.us.r-project.org")
if(!require(glue)) install.packages("glue", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
library(Peptides)
library(ggpubr)
library(broom)
library(glue)

# Load data and get a first overview
peptide_rt <- read_delim("unmod.txt", delim = '\t', show_col_types = FALSE)
head(peptide_rt)
str(peptide_rt)
```

Test of some functions from the Peptides package on the 20 natural amino acids.

```{r}
# Calculate the charge of amino acids at pH 2.7 (default pH for peptide chromatography)
sort(c(sapply(aaList(), function(x) charge(x, pH=2.7))))

# Calculate hydrophobicity for amino acids
sort(c(sapply(aaList(), hydrophobicity)))

# Calculate molecular weight
sort(c(sapply(aaList(), mw)))
```

```{r}
# Composition of amino acids in data set
aaComp(paste(peptide_rt$sequence, collapse = ""))
```

Based on the peptide sequence, different characteristics of the peptides are 
calculated and predicted.

```{r}
# Calculating and predicting peptide features
peptide_rt <- peptide_rt %>%
  mutate(length = str_length(sequence), # adding lenght of peptide
         charge = charge(sequence, pH=2.7), # adding charge
         hydrophobicity = hydrophobicity(sequence), # adding hydrophobicity
         molecular_weight = mw(sequence), # adding molecular weight
         aliphatic_index = aIndex(sequence), #adding aliphatic index
         )
head(peptide_rt)
```

Calculate the median peptide length

```{r}
median(peptide_rt$length)
```



The RT of peptides in minutes as well as the calculated/predicted
peptide features are plotted as histograms.
  
```{r}
pep_rt <- peptide_rt %>% 
  ggplot(aes(x = RT/60)) +
  geom_histogram(bins = 30) +
  ggtitle("Retention time (min)")

pep_length <- peptide_rt %>%
  ggplot(aes(x = length)) +
  geom_histogram(bins = 30) +
  ggtitle("Length")

pep_charge <- peptide_rt %>%
  ggplot(aes(x = charge)) +
  geom_histogram(bins = 12) +
  ggtitle("Charge")

pep_hydrophobicity <- peptide_rt %>%
  ggplot(aes(x = hydrophobicity)) +
  geom_histogram(bins = 30) +
  ggtitle("Hydrophobicity")

pep_mw <- peptide_rt %>%
  ggplot(aes(x = molecular_weight)) +
  geom_histogram(bins = 30) +
  ggtitle("Molecular weight")

pep_aliphatic <- peptide_rt %>%
  ggplot(aes(x = aliphatic_index)) +
  geom_histogram(bins = 30) +
  ggtitle("Aliphatic index")

ggarrange(pep_rt, pep_length, pep_charge, 
          pep_hydrophobicity, pep_mw, pep_aliphatic,
          ncol = 3, nrow = 2)
```

Scatter plots are used to show the relationships of RT to the different peptide
features.

```{r}
pep_rt_length <- peptide_rt %>% 
  ggplot(aes(x = RT/60, y = length)) +
  geom_point(alpha = 0.05) +
  ggtitle("RT vs. length") +
  geom_smooth(method = "loess")

pep_rt_charge <- peptide_rt %>% 
  ggplot(aes(x = RT/60, y = charge)) +
  geom_point(alpha = 0.05) +
  ggtitle("RT vs. charge") +
  geom_smooth(method = "loess")

pep_rt_hydrophobicity <- peptide_rt %>% 
  ggplot(aes(x = RT/60, y = hydrophobicity)) +
  geom_point(alpha = 0.05) +
  ggtitle("RT vs. hydrophobicity") +
  geom_smooth(method = "loess")

pep_rt_mw <- peptide_rt %>% 
  ggplot(aes(x = RT/60, y = molecular_weight)) +
  geom_point(alpha = 0.05) +
  ggtitle("RT vs. MW") +
  geom_smooth(method = "loess")

pep_rt_aliphatic <- peptide_rt %>% 
  ggplot(aes(x = RT/60, y = aliphatic_index)) +
  geom_point(alpha = 0.05) +
  ggtitle("RT vs. aliphatic index") +
  geom_smooth(method = "loess")

ggarrange(pep_rt_length, pep_rt_charge, pep_rt_hydrophobicity, 
          pep_rt_mw, pep_rt_aliphatic,
          ncol = 3, nrow = 2)
```



```{r}
# Split data in train and test set. Test set is 10% of peptide_rt data.
suppressWarnings(set.seed(1, sample.kind="Rounding"))
test_index <- createDataPartition(y = peptide_rt$RT, times = 1, p = 0.1, list = FALSE)
train_data <- peptide_rt[-test_index,]
dim(train_data)

test_data <- peptide_rt[test_index,]
dim(test_data)
```

Calculate the Root Mean Square Error (RMSE) of RT for linear models with different
numbers of peptide features.

```{r}
# Calculate the RMSE for predicting RT with peptide length
fit_l <- lm(RT ~ length, data = train_data)
RMSE(predict(fit_l, test_data), test_data$RT)/60

# Calculate the RMSE for predicting RT with peptide hydrophobicity
fit_h <- lm(RT ~ hydrophobicity, data = train_data)
RMSE(predict(fit_h, test_data), test_data$RT)/60

# Calculate the RMSE for predicting RT with peptide molecular weight
fit_m <- lm(RT ~ molecular_weight, data = train_data)
RMSE(predict(fit_m, test_data), test_data$RT)/60

# Calculate the RMSE for predicting RT with peptide aliphatic index
fit_a <- lm(RT ~ aliphatic_index, data = train_data)
RMSE(predict(fit_a, test_data), test_data$RT)/60

# Calculate the RMSE for predicting RT with peptide charge
fit_c <- lm(RT ~ charge, data = train_data)
RMSE(predict(fit_c, test_data), test_data$RT)/60

# Calculate the RMSE for predicting RT with all available peptide features
fit_lhmac <- lm(RT ~ length + hydrophobicity + molecular_weight + aliphatic_index + charge, data = train_data)
RMSE(predict(fit_lhmac, test_data), test_data$RT)/60
```

The linear model gave the smallest RMSE when all peptide features were included.
Thus, all peptide features will be used for testing additional models.

# 3. Results

The following models are evaluated to predict the RT of peptides in order to 
find the model that can best predict the RT with the smallest RMSE for the test 
data set:

- glm: Generalized Linear Model
- knn: k-Nearest Neighbour Classification
- svmLinear: linear support vector machines
- rf: random forest

In addition to calculating the RMSE for the test set, the time to train
the model will also be taken.

```{r}
# Define models to be tested and create empty vectors to record data
models <- c("glm", "knn", "svmLinear", "rf")
model_rmse <- numeric()
sec_per_model <- numeric()

# Train all models, take the time for each training and calculate RMSE on
# predictions for the test set
for (model in models){
  tic <- Sys.time()
  suppressWarnings(fit <- train(RT ~ length + hydrophobicity + 
    molecular_weight + aliphatic_index + charge,
	  method = model,
	  data = train_data))
  toc <- Sys.time()
	duration <- round(as.numeric(difftime(toc, tic, units="secs")),1)
	sec_per_model[model] <- duration
	
  rmse <- RMSE(predict(fit, test_data), test_data$RT)/60
  model_rmse[model] <- rmse
}

# Overview of time needed to train each model in seconds
sec_per_model

# Overview RMSE in min for RT predictions for the test set for each model
model_rmse

```


# 4. Conclusion

Four different models (glm: Generalized Linear Model, knn: k-Nearest Neighbour 
Classification, svmLinear: linear support vector machines, rf: random forest) 
were tested to predict the retention time of peptides based only on the amino acid
sequence and predicted/calculated physico-chemical properties of the peptides.
The rf model achieved the best results with the lowest RMSE of all four models.
However, the time to train the rf model was significantly longer (more than one hour)
compared to other models (only seconds). As seen in the exploratory data
analysis, there is a nearly linear relationship between the RT and some peptide
features. Therefore, it is no surprise that also the glm model gave a good result
(second smallest RMSE) with a very short time for training the model.
<br>
<br>
<br>
Andreas Hofmann
<br>
29-Dec-2021
<br>
_____

